---
title: "EDA"
author: "keimpe dijkstra"
date: "14-9-2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Importing used library's.
```{r}
library(ggplot2) 
```

Reading the data used for the project.
```{r}
data <- read.csv("./data/cirrhosis.csv")
```

# Cirrhosis prediction


## Introduction Cirrhosis

Cirrhosis is a serious dissease involving scarring, also called fibrosis, of the liver. 
The scarring develops due to sustained liver damage either by diseases like hepatitis B or C or by alcohol abuse.
Cirrhosis consists of four stages, the third of which is also referred to as cirrhosis.

### Stage 1
The first stage is often undiagnosed and is characterized by inflammation of the liver or bile duct, but this stage is critical because stage 1 is curable and so progression to stage 2 can be prevented.

### Stage 2
In the second stage the scarring and inflammation is starting to obstruct the the bloodflow within the liver.
The liver is not able to function properly but with treatment the liver may be able to recover slightly and further damage can be prevented.

### Stage 3
Cirrhosis occurs in the third stage, the liver is permanently scarred and thus scar tissue starts to replace healthy tissue causing it to become hard and lumpy. 
Blood is prevented to enter the liver and may leak into the spleen.

### Stage 4
In the fourth stage the liver starts to (acutely) fail and medical intervention is needed to prevent fatality.


## The dataset
The dataset consists of 424 Primary biliary cholangitis patients who where on a trial for D-penicillamine which was placebo controlled. 
312 of the instances contain complete data as these patients have participated the trial, the remaining individuals did not participate in the trial but consented to have basic measurements taken.

### Attributes
1. ID: unique identifier
2. N_Days: number of days between registration and the earlier of death, transplantation, or study analysis time in July 1986
3. Status: status of the patient C (censored), CL (censored due to liver tx), or D (death)
4. Drug: type of drug D-penicillamine or placebo
5. Age: age in [days]
6. Sex: M (male) or F (female)
7. Ascites: presence of ascites N (No) or Y (Yes)
  + The accumulation of fluid in the peritoneal cavity
8. Hepatomegaly: presence of hepatomegaly N (No) or Y (Yes)
  + An enlarged liver
9. Spiders: presence of spiders N (No) or Y (Yes)
10. Edema: presence of edema N (no edema and no diuretic therapy for edema), S (edema present without diuretics, or edema resolved by diuretics), or Y (edema despite diuretic therapy)
11. Bilirubin: serum bilirubin in [mg/dl]
12. Cholesterol: serum cholesterol in [mg/dl]
13. Albumin: albumin in [gm/dl]
14. Copper: urine copper in [ug/day]
15. Alk_Phos: alkaline phosphatase in [U/liter]
16. SGOT: SGOT in [U/ml]
17. Triglycerides: triglicerides in [mg/dl]
18. Platelets: platelets per cubic [ml/1000]
19. Prothrombin: prothrombin time in seconds [s]
20. Stage: histologic stage of disease (1, 2, 3, or 4)




```{r}
ggplot(data, aes(x=Stage)) +geom_bar()
```